---
title: "SUBMISSION FINAL"
output: html_document
date: "2023-04-01"
Student Number: 210345397
---
# Installing packages, Creating a working directory and database 
```{r installing packages}

# Connect to a SQLite database
library(RSQLite)
conn <- dbConnect(SQLite(), "airlinemain_r.db")

# Loading Packages package manager to make further package installation much more hassle free 
install.packages("pacman")
require(pacman)
detach("package:pacman", unload=TRUE) # For Base 

# Installing the tidyverse package to extend the capabilities of the base R
install.packages("tidyverse")
library(tidyverse)

# Install.packages("DBI")
library(DBI)

# Install.packages("dplyr")
library(dplyr)

# Preventing conflicts
library(conflicted)
conflict_prefer("filter", "dplyr") #code will make the dplr filter() and lag() functions the default functions used
conflict_prefer("lag", "dplyr")

```
```{r adding more things to import}
library(ggplot2)
install.packages("gridExtra")
```

```{r setting working directory}
setwd("/Users/mint/Downloads/SAVES☞/dataverse_files") 

getwd()

```
```{r Creation of database & tables}
conn <- dbConnect(RSQLite::SQLite(), "airlinemain_r.db")

# Loading in data from the csv files into 2 different variables which will be merged using the append function to form the ontime table.
ontime2003 <- read.csv("2003.csv.bz2", header = TRUE)
ontime2004 <- read.csv("2004.csv.bz2", header = TRUE)
ontime2005 <- read.csv("2005.csv.bz2", header = TRUE)
ontime2006 <- read.csv("2006.csv.bz2", header = TRUE)
ontime2007 <- read.csv("2007.csv.bz2", header = TRUE)

# Using dbWriteTable to create the ontime table with the help of the append argument to merge the 2 variables created above. We are also writing the table to the airlinemain_r.db database.
dbWriteTable(conn, "ontime", ontime2003, append = TRUE)
dbWriteTable(conn, "ontime", ontime2004, append = TRUE)
dbWriteTable(conn, "ontime", ontime2005, append = TRUE)
dbWriteTable(conn, "ontime", ontime2006, append = TRUE)
dbWriteTable(conn, "ontime", ontime2007, append = TRUE)

# Loading in data from the airports, carriers, & planes csv files & writing them to the arilinemain_r.db database.
airports <- read.csv("airports.csv", header = TRUE)
carriers <- read.csv("carriers.csv", header = TRUE)
planes <- read.csv("plane-data.csv", header = TRUE)

# Ensuring that other similiarly named files do not interfere with our clean file
dbWriteTable(conn, "airports", airports, overwrite = TRUE)
dbWriteTable(conn, "carriers", carriers, overwrite = TRUE)
dbWriteTable(conn, "planes", planes, overwrite = TRUE)

dbListTables(conn)

```
# Q1: When is the best time of day, day of the week, and time of year to fly to minimise delays? 

Step 1: Importing and loading required packages
Step 2: Creating a database to work in
Step 3: Data pre-processing; data-cleaning, data-trandformation, data aggregation
Step 4: Creating tables in our database
Step 5: Exploratory Data Analysis
Question 1.1: When is the best time of year to fly 
Answer 1.1: The best time of the year to fly is September
Step 6: Analysis of data by days using percentage and comparing this finding to the average minutes of delays for that day 
Question 1.2: When is the best day of the week to fly 
Answer 1.2: The best day of the week to fly is Saturday
Step 7: examining deviation of average airtime and elapse time over days of the week
Question 1.3: What is the best time of the day to fly?
Answer 1.3: The best time of the day to fly is from 0400 to 0559

```{r Exploratory Data Analysis}
# Connect to SQLite database
conn <- dbConnect(RSQLite::SQLite(), "airlinemain_r.db")

# Execute SQL query and fetch results
query1_by_year <- dbGetQuery(conn, "
  SELECT Year,
         AVG(Cancelled) * 100 AS `% of Cancelled flights`,
         AVG(Diverted) * 100 AS `% of Diverted flights`,
         (SUM(CASE WHEN ArrDelay > 0 THEN 1 ELSE 0 END) / CAST(COUNT(*) AS FLOAT)) * 100 AS `% of ArrDelayed flights`,
         (SUM(CASE WHEN DepDelay > 0 THEN 1 ELSE 0 END) / CAST(COUNT(*) AS FLOAT)) * 100 AS `% of DepDelayed flights`
  FROM ontime
  GROUP BY Year
  ORDER BY Year
")

# Display results in a table
query1_by_year %>%
  rename(`Year` = 1, `% of Cancelled flights` = 2, `% of Diverted flights` = 3, `% of ArrDelayed flights` = 4, `% of DepDelayed flights` = 5)
```
```{r Plotting variables by year}
Year_Names <- c("2003", "2004", "2005", "2006", "2007")
par(mfrow=c(1,4), mar=c(4,4,2,1), xpd=TRUE)  # Setting the layout and margins of the plot

# Creating a plot for the percentage of cancelled flights in each year from 2003 to 2007
barplot(query1_by_year[["% of Cancelled flights"]], xlab="Year", ylab="Cancelled flights %", main="Figure 1.1a: Cancelled flights % by Year", col="steelblue", ylim=c(0,3), names.arg=Year_Names)

# Creating a plot for the percentage of diverted flights in each year from 2003 to 2007
barplot(query1_by_year[["% of Diverted flights"]], xlab="Year", ylab="Diverted flights %", main="Figure 1.1b: Diverted flights % by Year", col="mediumseagreen", ylim=c(0,0.5), names.arg=Year_Names)

# Creating a plot for the percentage of arrival delayed flights in each year from 2003 to 2007
barplot(query1_by_year[["% of ArrDelayed flights"]], xlab="Year", ylab="ArrDelayed flights %", main="Figure 1.1c: ArrDelayed flights % by Year", col="mediumpurple", ylim=c(0,70), names.arg=Year_Names)

# Creating a plot for the percentage of departure delayed flights in each year from 2003 to 2007
barplot(query1_by_year[["% of DepDelayed flights"]], xlab="Year", ylab="DepDelayed flights %", main="Figure 1.1d: DepDelayed flights % by Year", col="mediumblue", ylim=c(0,50), names.arg=Year_Names)
```
From Figure 1.1, we observed that the percentage of overall cancellation, diversion, arrival delay and departure delay has increased each year. One deviation can be obeserved in 2006 where the percentage of cancelled flights was lower than in 2005

Further analysis is required on the individual variables and years to answer Q1
## Q1.1 When is the best time of the year to minimize delays
```{r Analysing data by month in 2006}
#Export dataframe to a SQlite database
conn <- dbConnect(RSQLite::SQLite(), dbname = "flight_data.sqlite")
DBI::dbWriteTable(conn, "ontime2006", ontime2006, append = TRUE)

#Execute a SQL query to calculate average cancellation, diversion, arrival delay, and departure delay percentages by month in 2006
Query2ByMonth <- dbGetQuery(conn, "SELECT Month, (AVG(Cancelled)*100), (AVG(Diverted)*100), ((CAST(SUM(CASE WHEN ArrDelay > 0 THEN 1 ELSE 0 END) AS FLOAT)/CAST(COUNT() AS FLOAT))*100), ((CAST(SUM(CASE WHEN DepDelay > 0 THEN 1 ELSE 0 END) AS FLOAT)/CAST(COUNT() AS FLOAT))*100)
FROM ontime2006
GROUP BY Month
ORDER BY Month")

#Rename the columns of the resulting data frame
names(Query2ByMonth) <- c("Month", "% of Cancelled flights", "% of Diverted flights", "% of ArrDelayed flights", "% of DepDelayed flights")

#View the resulting data frame
Query2ByMonth
```
```{r Plotting 2006 data by months according to}
    # "Cancelled flights %"
    # "Diverted flights %"
    # "ArrDelayed flights %"
    # "DepDelayed flights %"

month_names <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")

#Creating a 1x4 subplot figure
par(mfrow=c(1,4), mar=c(4,4,2,1), oma=c(0,0,2,0), mgp=c(2,0.7,0))

#Creating a plot for the percentage of cancelled flights in each month in 2006
barplot(Query2ByMonth[,2], names.arg = month_names, xlab = "Month", ylab = "Cancelled flights %",
main = "Figure 1.2a: Cancelled flights % by Month", col = "red", ylim = c(0,10), width = 0.6)

#Creating a plot for the percentage of diverted flights in each month in 2006
barplot(Query2ByMonth[,3], names.arg = month_names, xlab = "Month", ylab = "Diverted flights %",
main = "Figure 1.2b: Diverted flights % by Month", col = "mediumseagreen", ylim = c(0,1), width = 0.6)

#Creating a plot for the percentage of arrival delay flights in each month in 2006
barplot(Query2ByMonth[,4], names.arg = month_names, xlab = "Month", ylab = "ArrDelayed flights %",
main = "Figure 1.2c: Delayed flights % by Month", col = "mediumpurple", ylim = c(0,60), width = 0.6)

#Creating a plot for the percentage of departure delay flights in each month in 2006
barplot(Query2ByMonth[,5], names.arg = month_names, xlab = "Month", ylab = "DepDelayed flights %",
main = "Figure 1.2d: DepDelayed flights % by Month", col = "mediumblue", ylim = c(0,60), width = 0.6)
```
```{r Analysing data by month in 2007}
#Export dataframe to a SQlite database
conn <- dbConnect(RSQLite::SQLite(), dbname = "flight_data.sqlite")
DBI::dbWriteTable(conn, "ontime2007", ontime2007, append = TRUE)
#Execute a SQL query to calculate average cancellation, diversion, arrival delay, and departure delay percentages by month in 2007
Query3ByMonth <- dbGetQuery(conn, "SELECT Month, (AVG(Cancelled)*100), (AVG(Diverted)*100), ((CAST(SUM(CASE WHEN ArrDelay > 0 THEN 1 ELSE 0 END) AS FLOAT)/CAST(COUNT() AS FLOAT))*100), ((CAST(SUM(CASE WHEN DepDelay > 0 THEN 1 ELSE 0 END) AS FLOAT)/CAST(COUNT() AS FLOAT))*100)
FROM ontime2007
GROUP BY Month
ORDER BY Month")

#Rename the columns of the resulting data frame
names(Query3ByMonth) <- c("Month", "% of Cancelled flights", "% of Diverted flights", "% of ArrDelayed flights", "% of DepDelayed flights")

#View the resulting data frame
Query3ByMonth
```
```{r Plotting 2007 data by months }
Month_names <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")

par(mfrow=c(1,4), mar=c(4,4,2,1), oma=c(0,0,2,0), mgp=c(2,0.7,0))

# Creating a plot for the percentage of cancelled flights in each month in 2007 
barplot(Query3ByMonth[,"% of Cancelled flights"], names.arg=Month_names, main="Figure 1.3a: Cancelled flights % by Month",
        xlab="Month", ylab="Cancelled flights %", col="darkblue", border="white", ylim=c(0,10), cex.main=0.8)

# Creating a plot for the percentage of diverted flights in each month in 2007
barplot(Query3ByMonth[,"% of Diverted flights"], names.arg=Month_names, main="Figure 1.3b: Diverted flights % by Month",
        xlab="Month", ylab="Diverted flights %", col="darkgreen", border="white", ylim=c(0,1), cex.main=0.8)

# Creating a plot for the percentage of arrival delay flights in each month in 2007
barplot(Query3ByMonth[,"% of ArrDelayed flights"], names.arg=Month_names, main="Figure 1.3c: Delayed flights % by Month",
        xlab="Month", ylab="ArrDelayed flights %", col="darkmagenta", border="white", ylim=c(0,60), cex.main=0.8)

# Creating a plot for the percentage of departure delay flights in each month in 2007
barplot(Query3ByMonth[,"% of DepDelayed flights"], names.arg=Month_names, main="Figure 1.3d: DepDelayed flights % by Month",
        xlab="Month", ylab="DepDelayed flights %", col="darkred", border="white", ylim=c(0,60), cex.main=0.8)
```
```{r Analysing data of 2003-2007 by month}
#Execute a SQL query to calculate average cancellation, diversion, arrival delay, and departure delay percentages by month from 2003-2007
Query4ByMonth <- dbGetQuery(conn, "SELECT Month, (AVG(Cancelled)*100), (AVG(Diverted)*100), ((CAST(SUM(CASE WHEN ArrDelay > 0 THEN 1 ELSE 0 END) AS FLOAT)/CAST(COUNT() AS FLOAT))*100), ((CAST(SUM(CASE WHEN DepDelay > 0 THEN 1 ELSE 0 END) AS FLOAT)/CAST(COUNT() AS FLOAT))*100)
FROM ontime
GROUP BY Month
ORDER BY Month")

#Rename the columns of the resulting data frame
names(Query4ByMonth) <- c("Month", "% of Cancelled flights", "% of Diverted flights", "% of ArrDelayed flights", "% of DepDelayed flights")

#View the resulting data frame
Query4ByMonth
```
```{r Plotting 2003-2007 data by month}
Month_names <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")

par(mfrow=c(1,4), mar=c(4,4,2,1), oma=c(0,0,2,0), mgp=c(2,0.7,0))

# Creating a plot for the percentage of cancelled flights in each month in 2003-2007
barplot(Query4ByMonth[,"% of Cancelled flights"], names.arg=Month_names, main="Figure 1.4a: Cancelled flights % by Month",
        xlab="Month", ylab="Cancelled flights %", col="darkblue", border="white", ylim=c(0,10), cex.main=0.8)

# Creating a plot for the percentage of diverted flights in each month in 2003-2007
barplot(Query4ByMonth[,"% of Diverted flights"], names.arg=Month_names, main="Figure 1.4b: Diverted flights % by Month",
        xlab="Month", ylab="Diverted flights %", col="darkgreen", border="white", ylim=c(0,1), cex.main=0.8)

# Creating a plot for the percentage of arrival delay flights in each month in 2003-2007
barplot(Query4ByMonth[,"% of ArrDelayed flights"], names.arg=Month_names, main="Figure 1.4c: Delayed flights % by Month",
        xlab="Month", ylab="ArrDelayed flights %", col="darkmagenta", border="white", ylim=c(0,60), cex.main=0.8)

# Creating a plot for the percentage of departure delay flights in each month in 2003-2007
barplot(Query4ByMonth[,"% of DepDelayed flights"], names.arg=Month_names, main="Figure 1.4d: DepDelayed flights % by Month",
        xlab="Month", ylab="DepDelayed flights %", col="darkred", border="white", ylim=c(0,60), cex.main=0.8)
```
## A1.1 
Upon analysis of 2003-2007 data compared to 2007 and 2006, the results below remain true for 2006 and 2007 after data across all 5 years has been used. 

The best time of the year to minimise delays is September: 
The percentage of cancelled flights is lowest in May
The percentage of diverted flights is lowest in May
The percentage of arrival delay flights is lowest in September
The percentage of departure delay flights is lowest in September

## Q1.2 When is the best day of the week to fly to minimize delays? 
Analysis will be done based on the percentage of total delayed flights and the average minutes for delays flights accross days of the week.

```{r Plotting graphs for the percentage of delays accross days of the week}
# Execute the SQL query
Query6a <- dbGetQuery(conn, "
  SELECT DayOfWeek,
         (AVG(Cancelled)*100),
         (AVG(Diverted)*100),
         ((CAST(SUM(CASE WHEN ArrDelay > 0 THEN 1 ELSE 0 END) AS FLOAT)/CAST(COUNT(*) AS FLOAT))*100),
         ((CAST(SUM(CASE WHEN DepDelay > 0 THEN 1 ELSE 0 END) AS FLOAT)/CAST(COUNT(*) AS FLOAT))*100)
  FROM ontime
  GROUP BY DayOfWeek
  ORDER BY DayOfWeek
")

# Rename columns of the resulting data frame
names(Query6a) <- c("Day of Week",
                    "% of Cancelled flights",
                    "% of Diverted flights",
                    "% of ArrDelayed flights",
                    "% of DepDelayed flights")

# Define the days of the week
Day_of_week <- c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")

# Create the plots
par(mfrow=c(1,2))
# Percentage of ARRIVAL DELAY flights from 2003-2007
barplot(Query6a$`% of ArrDelayed flights`, names.arg = Day_of_week, main = "Figure 6.1: Arrival Delayed flights % by Day of Week", xlab = "Day of Week", ylab = "ArrDelayed Flights %", col = "mediumpurple")
# Percentage of DEPARTURE DELAY flights from 2003-2007
barplot(Query6a$`% of DepDelayed flights`, names.arg = Day_of_week, main = "Figure 6.2: Departure Delayed flights % by Day of Week", xlab = "Day of Week", ylab = "DepDelayed Flights %", col = "mediumblue")
```
```{r Plotting graphs for the ave minutes of delays accross days of the week}
# Query the database to extract information regarding average arrival delay by day of week from 2003-2007
Query6b <- dbGetQuery(conn, "
  SELECT DayOfWeek, AVG(ArrDelay), AVG(DepDelay)
  FROM ontime
  WHERE Cancelled = 0 AND Diverted = 0
  GROUP BY DayOfWeek
  ORDER BY DayOfWeek
")

# Rename columns of the resulting data frame
names(Query6b) <- c("Day of Week", "Avg Arrival Delay (mins)", "Avg Departure Delay (mins)")

# Define the days of the week
Day_of_week <- c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")

# Create a plot for the average arrival delay by day of week from 2003-2007
ggplot(Query6b, aes(x = `Day of Week`, y = `Avg Arrival Delay (mins)`)) +
  geom_bar(stat = "identity", fill = "firebrick", alpha = 1) +
  labs(x = "Day of Week", y = "Average Arrival Delay in Mins", 
       title = "Figure 6.3: Average Arrival Delay by Day of Week") +
  theme(axis.text.x = element_text(angle = 0, vjust = 0.5, hjust = 1))

# Create a plot for the average departure delay by day of week from 2003-2007
ggplot(Query6b, aes(x = `Day of Week`, y = `Avg Departure Delay (mins)`)) +
  geom_bar(stat = "identity", fill = "pink", alpha = 1) +
  labs(x = "Day of Week", y = "Average Departure Delay in Mins", 
       title = "Figure 6.4: Average Departure Delay by Day of Week") +
  theme(axis.text.x = element_text(angle = 0, vjust = 0.5, hjust = 1))
```
```{r examining deviation of average airtime and elapse time over days of the week}
#Determining the way time has been strucutred/ weeks from 2003-2007
Query7a <- dbGetQuery(conn, "
SELECT DayOfWeek, AVG(AirTime), AVG(ActualElapsedTime)
FROM ontime
GROUP by DayOfWeek
ORDER by DayOfWeek")

#Printing out query
names(Query7a) <- c("Day of Week", "Avg Airtime", "Avg Elaspse Time")
Query7a
```
## Q1.3: What is the best time of the day to fly to minimise delays?
```{r}
# Querying the database
Query7b <- dbSendQuery(conn,
                    "SELECT (AVG(Cancelled)*100), (AVG(Diverted)*100), ((CAST(SUM(CASE WHEN ArrDelay > 0 THEN 1 ELSE 0 END) AS FLOAT)/CAST(COUNT(*) AS FLOAT))*100),
                    CASE WHEN CRSDepTime BETWEEN 0000 AND 0159 THEN '0000 to 0159'
                         WHEN CRSDepTime BETWEEN 0200 AND 0359 THEN '0200 to 0359'
                         WHEN CRSDepTime BETWEEN 0400 AND 0559 THEN '0400 to 0559'
                         WHEN CRSDepTime BETWEEN 0600 AND 0759 THEN '0600 to 0759'
                         WHEN CRSDepTime BETWEEN 0800 AND 0959 THEN '0800 to 0959'
                         WHEN CRSDepTime BETWEEN 1000 AND 1159 THEN '1000 to 1159'
                         WHEN CRSDepTime BETWEEN 1200 AND 1359 THEN '1200 to 1359'
                         WHEN CRSDepTime BETWEEN 1400 AND 1559 THEN '1400 to 1559'
                         WHEN CRSDepTime BETWEEN 1600 AND 1759 THEN '1600 to 1759'
                         WHEN CRSDepTime BETWEEN 1800 AND 1959 THEN '1800 to 1959'
                         WHEN CRSDepTime BETWEEN 2000 AND 2159 THEN '2000 to 2159'
                         WHEN CRSDepTime BETWEEN 2200 AND 2359 THEN '2200 to 2359'
                         END AS TimeFrame
                    FROM ontime
                    GROUP BY CASE WHEN CRSDepTime BETWEEN 0000 AND 0159 THEN '0000 to 0159'
                                  WHEN CRSDepTime BETWEEN 0200 AND 0359 THEN '0200 to 0359'
                                  WHEN CRSDepTime BETWEEN 0400 AND 0559 THEN '0400 to 0559'
                                  WHEN CRSDepTime BETWEEN 0600 AND 0759 THEN '0600 to 0759'
                                  WHEN CRSDepTime BETWEEN 0800 AND 0959 THEN '0800 to 0959'
                                  WHEN CRSDepTime BETWEEN 1000 AND 1159 THEN '1000 to 1159'
                                  WHEN CRSDepTime BETWEEN 1200 AND 1359 THEN '1200 to 1359'
                                  WHEN CRSDepTime BETWEEN 1400 AND 1559 THEN '1400 to 1559'
                                  WHEN CRSDepTime BETWEEN 1600 AND 1759 THEN '1600 to 1759'
                                  WHEN CRSDepTime BETWEEN 1800 AND 1959 THEN '1800 to 1959'
                                  WHEN CRSDepTime BETWEEN 2000 AND 2159 THEN '2000 to 2159'
                                  WHEN CRSDepTime BETWEEN 2200 AND 2359 THEN '2200 to 2359'
                                  END
                    ORDER BY TimeFrame")
Query7b <- dbFetch(Query7b)
Query7b <- as.data.frame(Query7b)
Query7b <- Query7b[,c(4,1,2,3)]
names(Query7b) <- c("Time Interval", "% of Cancelled flights", "% of Diverted flights", "% of Delayed flights")
Query7b

```
## A1.3: The best time of the day to fly to minimize delays is from 0400 to 0559	as the delay percentage of the flight is the lowest at 28.87%. 

# ======================================================================
# Q2 Do older planes suffer more delays? 
We will be taking the following steps to analyse the plane age data to answer this question: 
Step 1: Querying the database to eliminate erroneous data
Step 3: Understanding the data we have under age of older planes
Step : Making the connection between the age of planes and delays
Step : Testing the strength in connection between age planes and delays
```{r Querying the database to identify the latest and earliest year of manufacture}
# Execute the SQL query and store the result in a data frame
Q2manufactureyearmax <- dbGetQuery(conn, "SELECT max(year) FROM planes WHERE year < 2100")

# Give a name to the column
names(Q2manufactureyearmax) <- "max_year"

# Rename the column in the data frame holding the query result
Q2manufactureyearmax <- dplyr::rename(Q2manufactureyearmax, `Latest Year of Manufacture` = max_year)

# Display the data frame
print(Q2manufactureyearmax)

# Execute the SQL query and store the result in a data frame
Q2manufactureyearmin <- dbGetQuery(conn, "SELECT min(year) FROM planes WHERE year > 1900")

# Give a name to the column
names(Q2manufactureyearmin) <- "min_year"

# Rename the column in the data frame holding the query result
Q2manufactureyearmin <- dplyr::rename(Q2manufactureyearmin, `Latest Year of Manufacture` = min_year)

# Display the data frame
print(Q2manufactureyearmin)
```
### Observations from finding the latest and the earliest year of manufacture: 
The latest year of manufacture is 2008 
The earliest year of manufacture is 1946
This totals to planes ranging from age 0 (considering we only include data till 2007) to age 61. We will be considering plane data with plane age 61 and above to be erroneous and emitting the data from our future analysis in this question.
```{r Removing erronous data}
# Removing erroneous data
# Any year of manufacture > year of flight = erroneous

# Query the database to find erroneous data
Q2erroneous <- dbGetQuery(conn, "SELECT ontime.Year, planes.year, (ontime.Year - planes.year) as Age_of_Plane
                                 FROM ontime JOIN planes ON ontime.TailNum = planes.tailnum
                                 WHERE (ontime.Year - planes.year) < 0 and (ontime.Year - planes.year) > 61
                                 ORDER BY (ontime.Year - planes.year)")

# Print the query result
Q2erroneous
```
# Querying the dataset for the distribution of non-erroneous data from 2003-2007
```{r Extracting non-erroneous data}
# Define SQL query
Q2total <- "SELECT ontime.Year, planes.year, (ontime.Year - planes.year)
              FROM ontime JOIN planes ON ontime.TailNum = planes.tailnum 
              ORDER BY (ontime.Year - planes.year)"

# Execute query and retrieve results as a data frame
Q2total <- dbGetQuery(conn, Q2total)
```
```{r Querying the non-erroneous database for plane age data}
Q2total <- dbGetQuery(conn, "SELECT ontime.Year, planes.year, (ontime.Year - planes.year)
                            FROM ontime JOIN planes ON ontime.TailNum = planes.tailnum 
                            ORDER BY (ontime.Year - planes.year)")

# Renaming columns of the dataframe holding the query result. 
names(Q2total) <- c("Year of flight", "Year of Manufacture", "Age of Plane")
Q2total
```
# Gaining visual on the plane age data
```{r Query for the spread of plane's age}
# Querying the database to extract information regarding the year of the plane manufacturing, age of the plane and year the plane was flown
Q2distribution <- dbGetQuery(conn, "SELECT ontime.Year, planes.year, (ontime.Year - planes.year)
                                    FROM ontime JOIN planes ON ontime.TailNum = planes.tailnum
                                    WHERE (ontime.Year - planes.year) > -1 AND (ontime.Year - planes.year) < 62
                                    ORDER BY (ontime.Year - planes.year)")

# Viewing the distribution queried in a table form
Q2distribution <- as.data.frame(Q2distribution)
# Renaming columns of the dataframe holding the query result. 
colnames(Q2distribution) <- c("Year of flight", "Year of Manufacture", "Age of Plane")
Q2distribution

# Plotting a histogram to observe the distribution of number of flights by the age of plane. 
ggplot(Q2distribution, aes(x = `Age of Plane`)) + 
  geom_histogram(bins = 62, colour = "black", fill = "white") + 
  labs(x = "Age of Plane", y = "Number of Flights in millions") + 
  ggtitle("Figure 2.1: Distribution of Number of Flights by Age of Plane") + 
  theme(plot.title = element_text(hjust = 0.5, size = 12))
```
### Observations from Figure 2.1:
The main age group of the planes range from ages 4 to 7 years old. The highest number of flights are flown by planes in this age group.
The number of flights made by plans under 20 years old is significantly higher than the number of flights made by planes that are more than 20 years old.
Hence, when doing our analysis it is important to keep in mind not to compare the delay of flights as a net number but to compare it as a ratio of the flights taken by planes in that particular age group.
Furthermore, due to a larger sample size of planes aged below 20, it may lead to differences in the accuracy of conclusion drawm from planes aged above 20 and below 20 years old.

```{r Plotting a jointplot of plane age against average ArrDelay}
# Querying the age of the plane against the variables that we are using to identify as delays
Q2planeageagainstvariables <- dbGetQuery(conn, "SELECT (ontime.Year - planes.year), AVG(ArrDelay), AVG(DepDelay), COUNT(*)
                                                FROM ontime JOIN planes ON ontime.TailNum = planes.tailnum
                                                WHERE (ontime.Year - planes.year) > -1 AND (ontime.Year - planes.year) < 62
                                                GROUP BY (ontime.Year - planes.year)
                                                ORDER BY (ontime.Year - planes.year) ASC")

# Renaming columns of the dataframe holding the query result.
colnames(Q2planeageagainstvariables) <- c("Age of Plane", "Average Arrival Delay in mins", "Average Departure Delay in mins", "Number of Flights")

# Plotting a scatterplot of the Ave ArrDelay against Age of Plane
scatterplot <- ggplot(Q2planeageagainstvariables, aes(x = `Age of Plane`, y = `Average Arrival Delay in mins`)) +
  geom_point(alpha = 0.4) + 
  geom_smooth(method = lm, se = FALSE) +
  labs(x = "Plane Age", y = "Average Arrival Delay in mins") +
  ggtitle("Figure 2.2.1: Scatterplot on Average Arrival Delay based on Age of Plane") +
  theme(plot.title = element_text(hjust = 0.5))

scatterplot
```

```{r Plotting a jointplot of plane age against average DepDelay}
# Convert data to data frame
Q2planeageagainstvariables <- as.data.frame(Q2planeageagainstvariables)

# Plotting a jointplot of the Ave DepDelay against Age of Plane
scatterplot <- ggplot(Q2planeageagainstvariables, aes(x = `Age of Plane`, y = `Average Departure Delay in mins`)) + 
               geom_point() + 
               geom_smooth(method = "lm") + 
               labs(x = "Plane Age", y = "Average Departure Delay in mins", 
                    title = "Figure 2.2.2: Scatterplot on Average Departure Delay based on Age of Plane") + 
               theme(plot.title = element_text(hjust = 0.5))

# Display the plot
scatterplot
```
### Observations from the Figure 2.2.1 and Figure 2.2.2: 
We can observe that there is no obvious pattern in linearity for both arrival and departure delay with respect to the age of the plane. From these charts however, we can deduce that higher number of flights made by planes between ages 0-10 does not imply increase number of delayed flights. Hence, we can analyse based on the fair and large quantity of data we have of planes ages 0-20 for the following analysis. 

## Querying the percentage of planes that are delayed by age
```{r extracting plane age relative to percentage of ArrDelay flights}
# Querying from database to extract information regarding the age of the planes relative to the percentage of its Arrival Delayed flights 
Q2delaycount <- dbGetQuery(conn, 
'SELECT (ontime.Year - planes.year), 
AVG(ArrDelay+DepDelay), 
COUNT(*), 
SUM(CASE WHEN ArrDelay+DepDelay > 0 THEN 1 ELSE 0 END), 
((CAST(SUM(CASE WHEN ArrDelay+DepDelay > 0 THEN 1 ELSE 0 END) AS FLOAT)/CAST(COUNT(*) AS FLOAT))*100) 
                           FROM ontime JOIN planes ON ontime.TailNum = planes.tailnum 
                           WHERE (ontime.Year - planes.year) > -1 AND (ontime.Year - planes.year) < 62 
                           GROUP BY (ontime.Year - planes.year) 
                           ORDER BY (ontime.Year - planes.year) ASC')

# Rename columns of the dataframe holding the query result
colnames(Q2delaycount) <- c("Age of Plane", "Average Arrival Delay in mins", "Total Number of Flights", "Number of Flights Delayed", "Percentage of Delayed Flights")

# Plotting a jointplot of the percentage of the Delayed Flights based on the Age of Plane
scatterplot <- ggplot(Q2delaycount, aes(x = `Age of Plane`, y = `Percentage of Delayed Flights`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Figure 2.3.1: Scatterplot on Percentage of Delayed Flights based on Age of Plane",
       x = "Plane Age", y = "Percentage of Delayed Flights") +
  theme_bw()

plot(scatterplot)
```
### Observations from the joint plot
The scatter plot shows the percentage of delayed flights based on the age of the plane. The regression of the plot above shows that an increase in the age of the plane is negatively correlated to the increase in the delay time of flights that the planes have departed on. 

As previously observed, the percentage of the delayed flights for planes older than 22 years old might not be as accurate due to the smaller sample size. However, by analysing the total percentages relative to flight delay. We can conclude that there is posititve and direct correlation between flight delay and the age of the plane.

### Further analysis: creating a subset for planes age 0-23 for a representative data set
```{r Plane age 0-23 and Delay count}
delaycountrepresentative <- dbGetQuery(conn, 
'SELECT (ontime.Year - planes.year), 
AVG(ArrDelay+DepDelay), 
COUNT(*), 
SUM(CASE WHEN ArrDelay+DepDelay > 0 THEN 1 ELSE 0 END), 
((CAST(SUM(CASE WHEN ArrDelay+DepDelay > 0 THEN 1 ELSE 0 END) AS FLOAT)/CAST(COUNT(*) AS FLOAT))*100) 
                           FROM ontime JOIN planes ON ontime.TailNum = planes.tailnum 
                           WHERE (ontime.Year - planes.year) > -1 AND (ontime.Year - planes.year) < 23
                           GROUP BY (ontime.Year - planes.year) 
                           ORDER BY (ontime.Year - planes.year) ASC')

# Rename columns of the dataframe holding the query result
colnames(delaycountrepresentative) <- c("Age of Plane", "Average Arrival Delay in mins", "Total Number of Flights", "Number of Flights Delayed", "Percentage of Delayed Flights")

# Plotting a jointplot to visualise the representation of the representative dataet
scatterplot <- ggplot(delaycountrepresentative, aes(x = `Age of Plane`, y = `Percentage of Delayed Flights`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Figure 2.3.2: Scatterplot on Percentage of Delayed Flights based on Age of Plane (Representative)",
       x = "Plane Age (Representative)", y = "Percentage of Delayed Flights") +
  theme_bw()

plot(scatterplot)
```
```{r Finding the R square and R value of the regression plot}
linear_model <- lm(`Percentage of Delayed Flights` ~ `Age of Plane`, data = delaycountrepresentative)
summary(linear_model)$r.squared

r_value <- sqrt(summary(linear_model)$r.squared)
r_value
```
## A.2: Older planes do suffer more delays
From Figure 2.3.2, we can conclude that plane age is strongly correlated (R= 0.88) to the percentage of delayed flights. However, we can only confirm this to be true for plane ages below 23 years of age as there is insufficient data for planes aged 23 and above to make any accurate analysis. 
# ======================================================================
# Q3.How does the number of people flying between different locations change over time?
```{r Finding the top 5 most travelled flight path (from one airport to another)}
Q3airport <- dbGetQuery(conn,
                 "SELECT ontime.*, 
                 ontime.Origin || ' to ' || ontime.Dest AS route,
                 COUNT(*) AS TotalFlights
                 FROM ontime 
                 JOIN airports 
                 ON ontime.dest = airports.iata
                 WHERE ontime.Cancelled = 0 AND ontime.Diverted = 0
                 GROUP by ontime.Origin, ontime.Dest
                 ORDER by COUNT(*) DESC"
                 )
# Creating a data frame from Q3airport
Q3airport_df <- data.frame(Q3airport)

# Renaming columns of Q3airport_df
colnames(Q3airport_df) <- c("Route", "Net Traffic")
Q3airport_df

```

```{r Plotting net traffic by year LAX to SAN}
Q3LAXtoSAN <- dbGetQuery(conn,
                 "SELECT ontime.Year, ontime.Month,
                 ontime.Origin || ' to ' || ontime.Dest AS route,
                 COUNT(*) AS TotalFlights
                 FROM ontime 
                 JOIN airports 
                 ON ontime.dest = airports.iata
                 WHERE ontime.Cancelled = 0 
                 AND ontime.Diverted = 0
                 AND route = 'LAX to SAN'
                 GROUP BY ontime.Year, ontime.Month, route
                 ORDER BY ontime.Year "
                 )

# Creating a data frame from Q3LAXtoSAN
Q3LAXtoSAN <- data.frame(Q3LAXtoSAN)

# Renaming columns of Q3LAXtoSAN
colnames(Q3LAXtoSAN) <- c("Year", "Month", "Route", "Net Traffic")

# Importing plotting libraries
library(ggplot2)
library(scales)
library(reshape2)

#PLOTTING NET TRAFFIC BY YEAR
# use ggplot2to find the route of the plane
x <- Q3LAXtoSAN$'Year'
y <- Q3LAXtoSAN$`Net Traffic`

ggplot(Q3LAXtoSAN, aes(x=x, y=y)) +
geom_line() +
labs(x="Year", y="Net Traffic", title="Net Traffic for LAX to SAN Route")

#LINEPLOT
ggplot(Q3LAXtoSAN, aes(x=Month, y=`Net Traffic`, colour=factor(Year))) +
geom_line() +
labs(x="Month", y="Net Traffic", title="Net Traffic for LAX to SAN Route") +
scale_colour_discrete(name="Year")
```
```{r SAN to LAX}
Q3SANtoLAX <- dbGetQuery(conn,
                 "SELECT ontime.Year, ontime.Month,
                 ontime.Origin || ' to ' || ontime.Dest AS route,
                 COUNT(*) AS TotalFlights
                 FROM ontime 
                 JOIN airports 
                 ON ontime.dest = airports.iata
                 WHERE ontime.Cancelled = 0 
                 AND ontime.Diverted = 0
                 AND route = 'SAN to LAX'
                 GROUP BY ontime.Year, ontime.Month, route
                 ORDER BY ontime.Year "
                 )

# Creating a data frame from Q3SANtoLAX
Q3SANtoLAX <- data.frame(Q3SANtoLAX)

# Renaming columns of Q3SANtoLAX
colnames(Q3SANtoLAX) <- c("Year", "Month", "Route", "Net Traffic")

# Importing plotting libraries
library(ggplot2)
library(scales)
library(reshape2)

#PLOTTING NET TRAFFIC BY YEAR
# use ggplot2to find the route of the plane
x <- Q3SANtoLAX$'Year'
y <- Q3SANtoLAX$`Net Traffic`

ggplot(Q3SANtoLAX, aes(x=x, y=y)) +
geom_line() +
labs(x="Year", y="Net Traffic", title="Net Traffic for SAN to LAX Route")

#LINEPLOT
ggplot(Q3SANtoLAX, aes(x=Month, y=`Net Traffic`, colour=factor(Year))) +
geom_line() +
labs(x="Month", y="Net Traffic", title="Net Traffic for SAN to LAX Route") +
scale_colour_discrete(name="Year")
```
```{r LAX to LAS}
Q3LAXtoLAS <- dbGetQuery(conn,
                 "SELECT ontime.Year, ontime.Month,
                 ontime.Origin || ' to ' || ontime.Dest AS route,
                 COUNT(*) AS TotalFlights
                 FROM ontime 
                 JOIN airports 
                 ON ontime.dest = airports.iata
                 WHERE ontime.Cancelled = 0 
                 AND ontime.Diverted = 0
                 AND route = 'LAX to LAS'
                 GROUP BY ontime.Year, ontime.Month, route
                 ORDER BY ontime.Year "
                 )

# Creating a data frame from Q3LAXtoLAS
Q3LAXtoLAS <- data.frame(Q3LAXtoLAS)

# Renaming columns of Q3LAXtoLAS
colnames(Q3LAXtoLAS) <- c("Year", "Month", "Route", "Net Traffic")

# Importing plotting libraries
library(ggplot2)
library(scales)
library(reshape2)

#PLOTTING NET TRAFFIC BY YEAR
# use ggplot2to find the route of the plane
x <- Q3LAXtoLAS$'Year'
y <- Q3LAXtoLAS$`Net Traffic`

ggplot(Q3LAXtoLAS, aes(x=x, y=y)) +
geom_line() +
labs(x="Time", y="Net Traffic", title="Net Traffic for LAX to LAS Route")

#LINEPLOT
ggplot(Q3LAXtoLAS, aes(x=Month, y=`Net Traffic`, colour=factor(Year))) +
geom_line() +
labs(x="Month", y="Net Traffic", title="Net Traffic for LAX to LAS Route") +
scale_colour_discrete(name="Year")
```
```{r LAX to LAS}
Q3BOStoLGA <- dbGetQuery(conn,
                 "SELECT ontime.Year, ontime.Month,
                 ontime.Origin || ' to ' || ontime.Dest AS route,
                 COUNT(*) AS TotalFlights
                 FROM ontime 
                 JOIN airports 
                 ON ontime.dest = airports.iata
                 WHERE ontime.Cancelled = 0 
                 AND ontime.Diverted = 0
                 AND route = 'BOS to LGA'
                 GROUP BY ontime.Year, ontime.Month, route
                 ORDER BY ontime.Year "
                 )

# Creating a data frame from Q3BOStoLGA
Q3BOStoLGA <- data.frame(Q3BOStoLGA)

# Renaming columns of Q3BOStoLGA
colnames(Q3BOStoLGA) <- c("Year", "Month", "Route", "Net Traffic")

# Importing plotting libraries
library(ggplot2)
library(scales)
library(reshape2)

#PLOTTING NET TRAFFIC BY YEAR
# use ggplot2to find the route of the plane
x <- Q3BOStoLGA$'Year'
y <- Q3BOStoLGA$`Net Traffic`

ggplot(Q3BOStoLGA, aes(x=x, y=y)) +
geom_line() +
labs(x="Time", y="Net Traffic", title="Net Traffic for BOS to LGA Route")

#LINEPLOT
ggplot(Q3BOStoLGA, aes(x=Month, y=`Net Traffic`, colour=factor(Year))) +
geom_line() +
labs(x="Month", y="Net Traffic", title="Net Traffic for BOS to LGA Route") +
scale_colour_discrete(name="Year")
```
## Querying top 3 routes by month to analyse difference in trends across months
```{r}
Q3Top3ByMonth <- dbGetQuery(conn,
                 "SELECT ontime.Year, ontime.Month,
                 ontime.Origin || ' to ' || ontime.Dest AS route,
                 COUNT(*) AS TotalFlights
                 FROM ontime 
                 JOIN airports 
                 ON ontime.dest = airports.iata
                 WHERE ontime.Cancelled = 0 
                 AND ontime.Diverted = 0
                 AND (Route = 'BOS to LGA'
                 OR Route = 'LAX to LAS'
                 OR Route = 'LAX to SAN')
                 GROUP BY ontime.Year, ontime.Month, route
                 ORDER BY ontime.Year ")

Q3Top3ByMonth <- data.frame(Q3Top3ByMonth)
colnames(Q3Top3ByMonth) <- c("Year", "Month", "Route", "Net Traffic")
Q3Top3ByMonth
```
```{r Plotting Net traffic for top 3 routes by month}
# plot the line graph with month on the x-axis and Net Traffic on the y-axis, colored by Route
ggplot(Q3Top3ByMonth, aes(x = Month, y = `Net Traffic`, color = Route)) +
  geom_line() +
  xlab("Month") +
  ylab("Net Traffic") +
  ggtitle("Net Traffic for Top 3 Routes against Month") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5))  # center the plot title

```

```{r Plotting net traffic for top 3 routes by year}
# Convert Year column to integer
Q3Top3ByMonth$Year <- as.integer(Q3Top3ByMonth$Year)

# Line plot
ggplot(Q3Top3ByMonth, aes(x = Year, y = `Net Traffic`, color = Route)) +
  geom_line() +
  labs(x = "Year", y = "Net Traffic", title = "Net Traffic for Top 3 Routes from 2003-2007") +
  scale_x_continuous(breaks = seq(2003, 2007, 1), labels = as.character(seq(2003, 2007, 1)))
```
From the tables above, the trends in the volume of air traffic differs from route to route, however it can be seen that the net traffic of the different routes changes over time. 
## A.3: We can observe in a change in the number of passengers over time according to their routes
# ======================================================================
# Q4. Can you detect cascading failures as delays in one airport create delays in others?
Query : Arrival delay of planes that have consecutive flights result in departure delay at the following locations
Query : Does the Departure delay that happen cascade to the next chain of airports? 

```{r Analysing top 5 delays under 100 mins}
# Query flights that have experienced "LateAircraftDelay < 100mins"
q4_lateaircraftdelay <- dbGetQuery(conn,
    "SELECT Year, Month, DayofMonth, DepTime, CRSDepTime, ArrTime, CRSArrTime, TailNum, ArrDelay, 
    DepDelay, Origin, Dest, FlightNum, LateAircraftDelay 
    FROM ontime 
    WHERE Cancelled = 0 AND Diverted = 0 AND LateAircraftDelay > 0 AND LateAircraftDelay < 100 
    AND TailNum != 0 
    ORDER BY LateAircraftDelay DESC"
)

# Rename columns
colnames(q4_lateaircraftdelay) <- c("Year", "Month", "DayofMonth", "DepTime", "CRSDepTime", "ArrTime", "CRSArrTime", "TailNum", "ArrDelay", "DepDelay", "Origin", "Dest", "FlightNum", "LateAircraftDelay")

q4_lateaircraftdelay
```
Top 5 rows 
2003	6	8	2127	1940	2225	2040	N341UA
2003	6	28	2038	1855	2202	2010	N926UA
2003	6	12	2400	2220	120	2341	N123UW
2003	6	23	1947	1750	2124	1911	N629AU
2003	6	13	2336	2000	220	2239	N178UW
```{r 8/6/2003 N341UA }
# Query flights with tail number N464UA on 8/6/2003 
q4_N464UA <- dbGetQuery(conn, "SELECT Year, Month, DayofMonth, DepTime, CRSDepTime, ArrTime, CRSArrTime, TailNum, ArrDelay,
                                DepDelay, Origin, Dest, LateAircraftDelay
                             FROM ontime
                             WHERE Year = 2003 AND Month = 6 AND DayofMonth = 8 AND TailNum = 'N341UA'
                             ORDER BY CRSDepTime")

# Rename columns
names(q4_N464UA) <- c("Year", "Month", "DayofMonth", "DepTime", "CRSDepTime", "ArrTime", "CRSArrTime", "TailNum", 
                      "ArrDelay", "DepDelay", "Origin", "Dest", "LateAircraftDelay")
q4_N464UA
```
```{r 8/6/2003 N341UA }
# Query flights with tail number N464UA on 8/6/2003 
q4_N464UA <- dbGetQuery(conn, "SELECT Year, Month, DayofMonth, DepTime, CRSDepTime, ArrTime, CRSArrTime, TailNum, ArrDelay,
                                DepDelay, Origin, Dest, LateAircraftDelay
                             FROM ontime
                             WHERE Year = 2003 AND Month = 6 AND DayofMonth = 8 AND TailNum = 'N341UA'
                             ORDER BY CRSDepTime")

# Rename columns
names(q4_N464UA) <- c("Year", "Month", "DayofMonth", "DepTime", "CRSDepTime", "ArrTime", "CRSArrTime", "TailNum", 
                      "ArrDelay", "DepDelay", "Origin", "Dest", "LateAircraftDelay")
q4_N464UA
```
```{r N926UA 6/28/2003}
# Query flights with tail number N926UA on 6/28/2003
q4_N926UA <- dbGetQuery(conn, "SELECT Year, Month, DayofMonth, DepTime, CRSDepTime, ArrTime, CRSArrTime, TailNum, ArrDelay,
                                DepDelay, Origin, Dest, LateAircraftDelay
                             FROM ontime
                             WHERE Year = 2003 AND Month = 6 AND DayofMonth = 28 AND TailNum = 'N926UA'
                             ORDER BY CRSDepTime")

# Rename columns
names(q4_N926UA) <- c("Year", "Month", "DayofMonth", "DepTime", "CRSDepTime", "ArrTime", "CRSArrTime", "TailNum", 
                      "ArrDelay", "DepDelay", "Origin", "Dest", "LateAircraftDelay")
q4_N926UA
```
```{r N123UW 6/12/2003}
# Query flights with tail number N123UW on 6/12/2003 
q4_N123UW <- dbGetQuery(conn, "SELECT Year, Month, DayofMonth, DepTime, CRSDepTime, ArrTime, CRSArrTime, TailNum, ArrDelay,
                                DepDelay, Origin, Dest, LateAircraftDelay
                             FROM ontime
                             WHERE Year = 2003 AND Month = 6 AND DayofMonth = 12 AND TailNum = 'N123UW'
                             ORDER BY CRSDepTime")

# Rename columns
names(q4_N123UW) <- c("Year", "Month", "DayofMonth", "DepTime", "CRSDepTime", "ArrTime", "CRSArrTime", "TailNum", 
                      "ArrDelay", "DepDelay", "Origin", "Dest", "LateAircraftDelay")
q4_N123UW
```
```{r N629AU 6/23/2003}
# Query flights with tail number N629AU on 6/23/2003 
q4_N629AU <- dbGetQuery(conn, "SELECT Year, Month, DayofMonth, DepTime, CRSDepTime, ArrTime, CRSArrTime, TailNum, ArrDelay,
                                DepDelay, Origin, Dest, LateAircraftDelay
                             FROM ontime
                             WHERE Year = 2003 AND Month = 6 AND DayofMonth = 23 AND TailNum = 'N629AU'
                             ORDER BY CRSDepTime")

# Rename columns
names(q4_N629AU) <- c("Year", "Month", "DayofMonth", "DepTime", "CRSDepTime", "ArrTime", "CRSArrTime", "TailNum", 
                      "ArrDelay", "DepDelay", "Origin", "Dest", "LateAircraftDelay")
q4_N629AU

```
```{r N178UW 6/13/2003}
# Query flights with tail number N178UW on 6/13/2003 
q4_N178UW <- dbGetQuery(conn, "SELECT Year, Month, DayofMonth, DepTime, CRSDepTime, ArrTime, CRSArrTime, TailNum, ArrDelay,
                                DepDelay, Origin, Dest, LateAircraftDelay
                             FROM ontime
                             WHERE Year = 2003 AND Month = 6 AND DayofMonth = 13 AND TailNum = 'N178UW'
                             ORDER BY CRSDepTime")

# Rename columns
names(q4_N178UW) <- c("Year", "Month", "DayofMonth", "DepTime", "CRSDepTime", "ArrTime", "CRSArrTime", "TailNum", 
                      "ArrDelay", "DepDelay", "Origin", "Dest", "LateAircraftDelay")
q4_N178UW
```
Conclusion: N123UW and N926UA out of 5 of the 99 min flights observed casecading delays. 
Further analysis is required to understand if cascading failures in one airport lead to another

### Using N926UA as an example flight to analyse further cascading delays
ORD	STL	
N926UA 6/28/2003
```{r N926UA 6/28/2003}
# Execute the query
q4_ORD <- dbGetQuery(conn, "
                         SELECT Year, Month, DayofMonth, DepTime, CRSDepTime, ArrTime, CRSArrTime, TailNum, ArrDelay,
                         DepDelay, Origin, Dest, LateAircraftDelay
                         FROM ontime
                         WHERE Year = 2003 AND Month = 6 AND DayofMonth = 28 AND Origin = 'ORD' AND LateAircraftDelay < 100 
                         ORDER BY CRSDepTime
                       ")

# Rename the columns
colnames(q4_ORD) <- c("Year", "Month", "DayofMonth", "DepTime", "CRSDepTime", "ArrTime", "CRSArrTime", "TailNum", "ArrDelay", 
                      "DepDelay", "Origin", "Dest", "LateAircraftDelay")
q4_ORD
```
```{r Analysing Dep and Arr delay of ORD to its following airports 2003}
# drop any rows that have missing values in either the ArrDelay or DepDelay column
install.packages("dplyr")
library(dplyr)
library(tidyr)
q4_ORD <- q4_ORD %>% drop_na(ArrDelay, DepDelay)

# select only the ArrDelay and DepDelay columns
arr_delay <- q4_ORD$ArrDelay
dep_delay <- q4_ORD$DepDelay

# calculate the correlation
correlation <- cor(arr_delay, dep_delay)

# print the correlation
print(correlation)
```

```{r 03/01/2004}
# Execute the query
q4_ORD2004 <- dbGetQuery(conn, "
                         SELECT Year, Month, DayofMonth, DepTime, CRSDepTime, ArrTime, CRSArrTime, TailNum, ArrDelay,
                         DepDelay, Origin, Dest, LateAircraftDelay
                         FROM ontime
                         WHERE Year = 2006 AND Month = 1 AND DayofMonth = 13 AND Origin = 'ORD' AND LateAircraftDelay < 100 AND Origin = 'ORD'
                         ORDER BY CRSDepTime
                       ")

# Rename the columns
colnames(q4_ORD2004) <- c("Year", "Month", "DayofMonth", "DepTime", "CRSDepTime", "ArrTime", "CRSArrTime", "TailNum", "ArrDelay", 
                      "DepDelay", "Origin", "Dest", "LateAircraftDelay")
q4_ORD2004
```
```{r Analysing Dep and Arr delay of ORD to its following airports 2004}
q4_ORD2004 <- q4_ORD2004 %>% drop_na(ArrDelay, DepDelay)

# select only the ArrDelay and DepDelay columns
arr_delay <- q4_ORD2004$ArrDelay
dep_delay <- q4_ORD2004$DepDelay

# calculate the correlation
correlation <- cor(arr_delay, dep_delay)

# print the correlation
print(correlation)
```
A.4: Cascading failures from one airport to another can be detected via analysis of arrival and departure delays in that airport. In this study, we tracked a late aircraft tailnumber and its late arrival into ORD airport. This late arrival caused a following late departure of other flights in that airport for Corr = 0.83 in 2003 and Corr = 0.84 in 2004. We can conclude that one delay is strongly correlated to further delays in the following airports.

# ======================================================================

# Q5. Use the available variables to construct a model that predicts delays
```{r}
install.packages("paradox")
```
```{r Importing relevant machine learning packages}
install.packages("mlr3")
install.packages("mlr3learners")
install.packages("mlr3pipelines")
install.packages("mlr3tuning")
install.packages("mlr_pipeops")

library(mlr3)
library(mlr3learners)
library(mlr3pipelines)
library(mlr3tuning)
library(mlr_pipeops)
```
```{r importing dataframe from python}
#Q5n_5000 <-read.csv("Q5n_5000.csv", header=TRUE) 
```
```{r Assigning  n sample row value}
#n <- nrow(q5new) 
```
```{r}
# Import required packages
library(mlr3)
library(mlr3learners)
library(mlr3pipelines)
library(mlr3tuning)
library(mlr_pipeops)

# Import dataset
Q5_DelaybyVariables <- dbGetQuery(conn, 
"SELECT Year, Month, DayofMonth, DayOfWeek, ArrTime, DepTime, Distance, ActualElapsedTime, AirTime, CarrierDelay, WeatherDelay, NASDelay, SecurityDelay, ArrDelay, DepDelay, LateAircraftDelay
FROM ontime
WHERE Cancelled = 0 AND Diverted = 0 AND LateAircraftDelay > 0
AND TailNum != 0
ORDER BY LateAircraftDelay DESC")

# Split data into training and test sets
set.seed(123)
train_index <- sample(nrow(Q5_DelaybyVariables), 0.7 * nrow(Q5_DelaybyVariables))
train <- Q5_DelaybyVariables[train_index,]
test <- Q5_DelaybyVariables[-train_index,]

# Train models
lmModelad <- lm(ArrDelay ~ ., data=train)
lmModeldd <- lm(DepDelay ~ ., data=train)

# Predict on test set
test$PredictedArrDelay <- predict(lmModelad, test)
test$PredictedDepDelay <- predict(lmModeldd, test)

# Calculate accuracy
acc_ad <- 1 - mean((test$ArrDelay - test$PredictedArrDelay) ^ 2) / var(test$ArrDelay)
acc_dd <- 1 - mean((test$DepDelay - test$PredictedDepDelay) ^ 2) / var(test$DepDelay)

# Print results
cat("Test dataset accuracy is for ArrDelay:", round(acc_ad, 2), "\n")
cat("Test dataset accuracy is for DepDelay:", round(acc_dd, 2), "\n")

```
A5: The best model for predicting late aircraft delay is the Linear Regression model, as shown from the results above. The linear regression model has a high level of accuracy and a low value of error
```{r Disconnect from database}

dbDisconnect(conn)

```